### 2011-01-04 Documentation/atropos/atropos.tex
ENH: this is the actual version submitted first round to neuroinformatics
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@735 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\newcommand{\lstsetcpp}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
       numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt Atropos short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides additional details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.},
        captionpos=b,
        label=listing:command
        }
}

\newcommand{\lstsetcppnfour}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt N4 short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides additional details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.  },
        captionpos=b,
        label=listing:n4
        }
}

@@

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and further subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended non-parametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various techniques have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \citep{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, including Klein's
Mindboggle \citep{Klein2005}, would usually follow an initial segmentation by a 
volumetric method such as Atropos.

@@

We also allow Atropos to address the bias field issue by incorporating
the recently developed N4 bias correction software \citep{Tustison2010} in an adaptive manner.  Coupled with the
registration \citep{Avants2011} and template building 
\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
versatile and powerful software tool which touches multiple aspects of
our brain processing pipeline from brain extraction
\citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid
segmentation to label fusion and propagation for achieving a refined
cortical parcellation based on both prior information and
subject-specific image data.

%Atropos is multi-platform, free, /stand-alone and able to
%manage a wide range of scalar and vector image datatypes.

To highlight the
value of this open source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
%Atropos's memory efficiency enables EM-segmentation of many classes at
%once, a capability that is due to a sparse internal representation of
%prior probability images that minimizes the need for random access
%memory.  
We evaluate EM-based classification of 1 mm$^3$ T1 MR
neuroimages into 69 neuroanatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  
Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while
acknowledging that additional theoretical discussion is available elsewhere.  
This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open source contribution in the context of the remainder of
this paper and of previous and future work.

@@

Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.  The theory underlying Atropos dates back 20$+$ years and is representative of some of the most innovative work in the field.  Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.

@@

We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., 
 \begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x} 
 \end{align}
and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is then, 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ is the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

@@

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user-input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alterative to Boykov's solution.
Furthermore, one may automate the initialization process.  We compare
our sparse prior labeling strategy to fully automated methods in the
evaluation section.

@@

each iteration as in \cite{Ashburner2005}.%

@@

prior probability images, one for each class.

@@

\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i)

@@

\section{Implementation}

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=160mm]{AtroposFlowchart.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction via N4.  
Initialization provides an estimate before the iterative optimization in which the 
likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
posterior probabilities associated with each class.  The multiple options associated with the different algorithmic 
components are indicated by the colored rounded rectangles connected to their respective core Atropos 
processes via curved dashed lines.  }
\label{fig:flowchart}
\end{center}
\end{figure}
\subsection{The Atropos User Interface}
As with other classes that comprise ANTs, Atropos uses the Insight
Toolkit as a developmental foundation.  This allows us to take
advantage of the mature portions of ITK (e.g. image IO) and ensures
the integrity of the ancillary processes such as those facilitated by
the underlying statistical framework.  Although Atropos is publicly
distributed with the rest of the ANTs package, we plan to contribute
its core elements to the Insight Toolkit where it can be vetted and further
improved by other interested researchers.

@@

\doublespacing
\subsection{Initializing the Atropos Objective}
Atropos has a number of parameters defined within
Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
data.  However, the majority of the time, users will be concerned with
a smaller set of input parameters.  Here, we list the recommended input and
an example definition for each parameter:
\begin{description}
\item[Input images to be segmented:] If more than
  one input image is passed, then a multivariate model will be
  instantiated.  E.g.  {\ttfamily -a Image.nii.gz} for one image and
  {\ttfamily -a Image1.nii.gz -a Image2.nii.gz} for multiple images.
\item[Input image mask:] This binary image
  defines the spatial segmentation domain.  Voxels outside the
  masked region are designated with the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
\item[Convergence criteria:] The algorithm terminates if it reaches  the maximum number of iterations or
  produces a change less than the minimum threshold change in the posterior.
  E.g. {\ttfamily -c [5,1.e-5]}. 
\item[MRF prior:] The key parameter to increase or decrease the
  spatial smoothness of the label map is $\beta$.  A useful range of
  $\beta$ values is $0$ to $0.5$ where we usually use 0.05, 0.1 or 0.2
  in brain segmentation.  E.g.  {\ttfamily -m [0.1, 1x1x1]} would
  define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
\item[Initialization:] The initialization options include (where the
  first parameter defines $K$, here 3 for each below),
\begin{itemize}
\item  {\ttfamily -i  Kmeans[3]} standard K-means initialization for
  three classes,
\item  {\ttfamily -i  PriorLabelImage[3,label\_image.nii.gz]} and
%  where scalar $w \in [0,1]$ is unused. 
\item  {\ttfamily -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,w]} 
where $w=0$~(use the prior probability images only for
  initialization) or $w>0.0$ (use the prior probability images
  throughout the optimization).  If one chooses $0 < w < 1.0$ then one
  will increase (from zero) the weight on the priors.  These images,
  like the {\ttfamily PriorLabelImage}, should be defined with the same domain as
  the input images to be segmented.  
\end{itemize}
\item[Posterior formulation:]  The user may choose to estimate the
  mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
  selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
\item[Output:] Atropos will output the hard segmentation and the
  probability image for each model.  E.g.  {\ttfamily -o
    [segmentation.nii.gz,seg\_prob\%02d.nii.gz]} will write
  out the hard segmentation in the first output parameter and a
  probability image for each class named, here, seg\_prob01.nii.gz,
  seg\_prob02.nii.gz, etc. 
\end{description}
Higher dimensions than 4 are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same
dimension, size, origin, etc.), will automatically enable multivariate
likelihoods.  In that case, the first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.    More details on each of
the key implementation options are given below.

@@

\subsection{Prior Probability Models}
Consistent with our previous discussion, we offer both an MRF-based
prior probability for modeling spatial coherence and the possibility
of specifying a set of prior probability maps or a prior label map
with the latter extendable to creating a dense labeling.  To invoke
the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
factor (or the granularity parameter, $\beta$, given in
Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
system using the vector notation `{\ttfamily 1x1x1}' for a
neighborhood radius of 1 in all 3 dimensions.  This radius is defined
such that voxels including but not limited to those that are face
connected will influence the MRF.

\begin{comment}{
The user also has an option to propagate initial labels, set by the
prior options, by a distance map tool.  
To propagate labels with the specification of prior probability images
or a prior label image, one needs to select the distance function (set
by the `{\ttfamily -e/--use-euclidean-distance}' boolean option)
which, by default, is determined by the geodesic distance.  The user
also needs to set the label propagation parameters for one or more of
the classes.   This  includes both the boundary probability and the exponential
 decay parameter, respectively $\alpha_k$ and $\sigma_k$ in Eqn. (\ref{eq:prop}).  
}\end{comment}

>>>>

\newcommand{\lstsetcpp}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
       numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt Atropos short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.},
        captionpos=b,
        label=listing:command
        }
}

\newcommand{\lstsetcppnfour}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt N4 short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.  },
        captionpos=b,
        label=listing:n4
        }
}

@@

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended non-parametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Various techniques
have been proposed to reduce the manual effort required to densely
label a high-resolution neuroimage; one example is the popular
software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces \citep{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, including Klein's
Mindboggle \citep{Klein2005}, would usually follow an initial
segmentation by a volumetric method such as Atropos.

@@

Coupled with the registration \citep{Avants2011} and template building
\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
versatile and powerful software tool which touches multiple aspects of
our brain processing pipeline.  We use Atropos to address brain
extraction \citep{Avants2010a}, grey matter/white matter/cerebrospinal
fluid segmentation, label fusion/propagation and cortical
parcellation.  We also allow Atropos to interact with the recently
developed N4 bias correction software \citep{Tustison2010} in an
adaptive manner.  To further highlight the value of this open source
contribution, we performed a search of software attributes on NITRC
and found that as of November 2010 no stand-alone EM methods are
currently listed.  We also evaluate Atropos performance on two brain
MRI segmentation objectives.  First, three tissue classification.
Second, we test our ability to parcellate the brain into 69
neuroanatomical regions to illustrate the practical value of the
low-memory implementation within this paper.  Although Atropos may be
applied to multivariate data from arbitrary modalities, we limit our
evaluation to tissue-classification in T1 neuroimaging in part due to
the abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while
acknowledging that more theoretical discussion is available elsewhere.  
This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open source contribution in the context of the remainder of
this paper and of previous and future work.

@@

Atropos encodes a family of Bayesian segmentation techniques that may
be configured in an application-specific manner.  The theory
underlying Atropos dates back 20$+$ years and is representative of
some of the most innovative work in the field.  Although we synthesize
some of the theoretical work in this section, we recommend to the
interested reader consultation with the literature cited previously
for additional perspective.

@@

We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., 
 \begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x} 
 \end{align}
and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is then, 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ is the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more
heavily than distant sites.

@@

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user-input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alternative to Boykov's solution.
Furthermore, one may automate the initialization process.  We compare
our sparse prior labeling strategy to fully automated methods in the
evaluation section.

@@

each iteration, similar to \cite{Ashburner2005}.%

@@

prior probability images, one for each class.  The initialization also
provides starter parameters.

@@

\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i).

@@

\section{Implementation} 
Organization of the implementation section roughly follows that of the
theory section.  
%Thus, the reader can refer to, for instance, 2.2 for
%the likelihood theory and 3.3 for likelihood implementation.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=160mm]{AtroposFlowchart.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction via N4.  
Initialization provides an estimate before the iterative optimization in which the 
likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
posterior probabilities associated with each class.  The multiple options associated with the different algorithmic 
components are indicated by the colored rounded rectangles connected to their respective core Atropos 
processes via curved dashed lines.  }
\label{fig:flowchart}
\end{center}
\end{figure}
\subsection{The Atropos User Interface}
As with other classes that comprise ANTs, Atropos uses the Insight
Toolkit as a developmental foundation.  This allows us to take
advantage of the mature portions of ITK (e.g. image IO) and ensures
the integrity of the ancillary processes such as those facilitated by
the underlying statistical framework.  Although Atropos is publicly
distributed with the rest of the ANTs package, we plan to contribute
its core elements to the Insight Toolkit where it can be vetted and improved by other interested researchers.

@@

\doublespacing
\subsection{Initializing the Atropos Objective}
Atropos has a number of parameters defined within
Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
data.  However, the majority of the time, users will be concerned with
a smaller set of input parameters.  Here, we list the recommended input and
an example definition for each parameter:
\begin{description}
\item[Input images to be segmented:] If more than
  one input image is passed, then a multivariate model will be
  instantiated.  E.g.  {\ttfamily -a Image.nii.gz} for one image and
  {\ttfamily -a Image1.nii.gz -a Image2.nii.gz} for multiple images.
\item[Input image mask:] This binary image
  defines the spatial segmentation domain.  Voxels outside the
  masked region are designated with the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
\item[Convergence criteria:] The algorithm terminates if it reaches  the maximum number of iterations or
  produces a change less than the minimum threshold change in the posterior.
  E.g. {\ttfamily -c [5,1.e-5]}. 
\item[MRF prior:] The key parameter to increase or decrease the
  spatial smoothness of the label map is $\beta$.  A useful range of
  $\beta$ values is $0$ to $0.5$ where we usually use 0.05, 0.1 or 0.2
  in brain segmentation.  E.g.  {\ttfamily -m [0.1, 1x1x1]} would
  define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
\item[Initialization:] The initialization options include (where the
  first parameter defines $K$, here 3 for each below),
\begin{itemize}
\item  {\ttfamily -i  Kmeans[3]} standard K-means initialization for
  three classes,
\item  {\ttfamily -i  PriorLabelImage[3,label\_image.nii.gz]} and
%  where scalar $w \in [0,1]$ is unused. 
\item  {\ttfamily -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,w]} 
where $w=0$~(use the prior probability images only for
  initialization) or $w>0.0$ (use the prior probability images
  throughout the optimization).  If one chooses $0 < w < 1.0$ then one
  will increase (from zero) the weight on the priors.  These images,
  like the {\ttfamily PriorLabelImage}, should be defined with the same domain as
  the input images to be segmented.  
\end{itemize}
\item[Posterior formulation:]  The user may choose to estimate the
  mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
  selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
\item[Output:] Atropos will output the hard segmentation and the
  probability image for each model.  E.g.  {\ttfamily -o
    [segmentation.nii.gz,seg\_prob\%02d.nii.gz]} will write
  out the hard segmentation in the first output parameter and a
  probability image for each class named, here, seg\_prob01.nii.gz,
  seg\_prob02.nii.gz, etc. 
\end{description}
Higher dimensions than 4 are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same
dimension, size, origin, etc.), will automatically enable multivariate
likelihoods.  In that case, the first image specified on the command
line is used to initialize the Random, Otsu, or K-means labeling with
the latter initialization refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is
determined from the first intensity image which, along with the other
images, provides the starting multivariate cluster centers for a
follow-up multivariate K-means labeling.  More details on each of the
key implementation options are given below.

@@

\subsection{Prior Probability Models}
\subsubsection{Label Regularity}
Consistent with our previous discussion, we offer both an MRF-based
prior probability for modeling spatial coherence and the possibility
of specifying a set of prior probability maps or a prior label map
with the latter extendable to creating a dense labeling.  To invoke
the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
factor (or the granularity parameter, $\beta$, given in
Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
system using the vector notation `{\ttfamily 1x1x1}' for a
neighborhood radius of 1 in all 3 dimensions.  This radius is defined
such that voxels including but not limited to those that are face
connected will influence the MRF.
